Normalization by Evaluation for Sized Dependent Types

### 2017-02-21 nbe/icfp17.tex
[ ICFP 17 ] Working on "Into the LR".

\fbox{$\Gamma \der a : \Size$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der a : \Size$).
\[
  \ru{\der \Gamma}{\Gamma \der \infty : \Size}
\qquad
  \ru{\der \Gamma \qquad \Gamma(i) = \erann[\noterased]\Size
    }{\Gamma \der \ind i + o : \Size
    }
\]

@@

      \Gamma \der a : \Size \qquad
      \resurrect \Gamma \der b : \Size

@@

  \ru{\resurrect\Gamma \der b : \Size \qquad

@@

      \Gamma \der a : \Size \qquad

@@

    }{\Gamma \der (\zero{a})\,\case[\ell] T {t_z} {t_s} = t_z : T\,\zero{a_z}}

@@

  \ru{\Gamma \der t : \Nat b \qquad

@@

      \Gamma \der a : \Size \qquad

@@

    }{\Gamma \der (\suc{a} t)\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,(\suc{a} t)}

@@

      \Gamma \der a : \Size \qquad

@@

  \ru{\Gamma \der t : \forallT T \qquad \erext \Gamma \Size \der a : \Size

@@

      \Gamma \der a, a' : \Size \qquad
      \resurrect\Gamma \der b : \Size

@@

  \ru{\Gamma \der a, a' : \Size \qquad
      \resurrect\Gamma \der b : \Size

@@

  \ru{\Gamma \der a, a' : \Size \qquad

@@

Substitution typing. % and equality.
\fbox{$\Gamma \der \tau : \Delta$}
(implies $\der \Gamma$ and $\der \Delta$
 [and $\resurrect\Gamma \der \tau : \Delta$]
 and $\resurrect\Gamma \der \tau : \resurrect \Delta$).

\begin{gather*}
  \ru{\der \Gamma \qquad
    }{\Gamma \der \sempty : \cempty}
\qquad
  \ru{\Gamma \der \tau : \Delta \qquad
      \resurrect \Delta \der T \qquad
      \Gamma \der t : T \tau
    }{\Gamma \der \sext \tau t : \cext \Delta T}
\qquad
  \ru{\Gamma \der \tau : \Delta \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \sext \tau a : \eext \Delta \Size}
\end{gather*}

Weakenings \fbox{$\xi : \Gamma \leq \Delta$}  (implies $\der \Gamma$ and $\der \Delta$ and $\Gamma \der \xi : \Delta$).
\begin{gather*}
  \ru{\der \Gamma
    }{\sempty : \Gamma \leq \cempty}
\qquad
  \ru{\xi : \Gamma \leq \Delta \qquad
      \resurrect\Gamma \der T
    }{\xi \slift : \cext \Gamma T \leq \Delta}
\qquad
  \ru{\xi : \Gamma \leq \Delta \qquad
      \resurrect \Delta \der U \qquad
      \resurrect\Gamma \der T \leq U \xi
    }{(\xi \slift, \ind 0) : \cext \Gamma T \leq \cext \Delta U}
\end{gather*}

@@

\begin{gather*}
  \ru{\der \Gamma \qquad
    }{\sueq \Gamma \sempty \sempty \sempty \cempty}
\qquad
  \ru{\sueq \Gamma \sigma {\sigma'} \tau \Delta \qquad
      \resurrect \Delta \der T \qquad
      \Gamma \der u = u' = t : T \tau
    }{\sueq \Gamma {\sext \sigma u } {\sext {\sigma'} {u'}} {\sext \tau t} {\cext \Delta T}}
\\[2ex]
  \ru{\sueq \Gamma \sigma {\sigma'} \tau \Delta \qquad
      \Gamma \der a = a' = b : T \tau
    }{\sueq \Gamma {\sext \sigma a } {\sext {\sigma'} {a'}} {\sext \tau b} {\cext \Delta \Size}}
\qquad
  \ru{\sueq \Gamma \sigma {\sigma'} \tau \Delta \qquad
      \Gamma \der a, a', b : T \tau
    }{\sueq \Gamma {\sext \sigma a } {\sext {\sigma'} {a'}} {\sext \tau b} {\erext \Delta \Size}}
\end{gather*}

@@

\begin{lemma}[Resurrection]
\label{lem:res} \bla
\begin{enumerate}
\item $\der \Gamma$ iff $\der \resurrect\Gamma$;  then $\resurrect \Gamma \der \sid : \Gamma$.
  % Consequently, if\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
  % Further, if $\Gamma \der J : \resurrect\Delta$ then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
\item If\/ $\Gamma \der J : \resurrect \Delta$ % and $\der \Delta$
  then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der \sigma : \Delta$ then $\Gamma \der \sigma : \resurrect\Delta$.
% then $\resurrect\Gamma \der \sigma : \resurrect\Delta$.
\end{enumerate}
\end{lemma}

@@

\begin{lemma}[Substitution equality]
\label{lem:subeq}\bla
\begin{enumerate}
\item Conversion: If\/ $\sueq \Gamma \sigma {\sigma'} {\tau_1} \Delta$ and $\sueq {\resurrect\Gamma} {\tau_1} {\tau_2} {\tau} {\resurrect\Delta}$ then $\sueq \Gamma \sigma {\sigma'} {\tau_2} \Delta$.
\item Reflexivity: If\/ $\Gamma \der \sigma : \Delta$ then $\sueq \Gamma \sigma \sigma \sigma \Delta$.
\item Symmetry:  If\/ $\sueq \Gamma \sigma {\sigma'} {\tau} \Delta$ then $\sueq \Gamma {\sigma'} {\sigma} {\tau} \Delta$.
\item Transitivity:  If\/ $\sueq \Gamma {\sigma_1} {\sigma_2} {\tau} \Delta$ and $\sueq \Gamma {\sigma_2} {\sigma_3} {\tau} \Delta$ then $\sueq \Gamma {\sigma_1} {\sigma_3} {\tau} \Delta$.
\item Functionality: Let $\sueq \Gamma \sigma {\sigma'} \tau \Delta$.
  \begin{enumerate}
  \item
  If $\Delta \der t : T$ then $\Gamma \der t\sigma = t\sigma' : T\tau$.
  \item
  If $\Delta \der t = t' : T$ then $\Gamma \der t\sigma = t'\sigma' : T\tau$.
  \item
  Corollary: If $\Delta \der T \leq T'$ then $\Gamma \der T\sigma \leq T\sigma'$.
  \end{enumerate}
\end{enumerate}
\end{lemma}

@@

\begin{theorem}[Reflection and reification]
Let $T \in \SET[\ell]$ and $T \shape S_1$ and $T \shape S_2$.
\begin{enumerate}
\item If $n_1 = n_2 \in \NE$ then $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$.
\item If $t_1 = t_2 \in T$ then $\down{S_1}{t_1} = \down{S_2}{t_2} \in \NF$.
\end{enumerate}
\end{theorem}
\begin{proof}
By induction on $T \in \SET[\ell]$ and cases on $T \shape S_1$ and $T \shape S_2$.
\begin{caselist}

@@

\begin{corollary}
  If\/ $T \in \SET[\ell]$ and $t = t' \in T$ then $\down T t = \down T {t'} \in \NF$.
\end{corollary}

@@

\begin{theorem}[Fundamental theorem]
  \bla
  \begin{enumerate}
  \item
  If $\der \Gamma$ then $\models \Gamma$.
  \item
  If $\Gamma \der J$ then % $\models \Gamma$ and
  $\Gamma \models J$
  \end{enumerate}
\end{theorem}
\begin{proof}
Simultaneously, by induction on the derivation.
% The soundness of subtyping and the conversion rule is due to subsumption (lemmata \ref{lem:subsump} and \ref{lem:usubsump}).

@@

@@

\[
  \fbox{$\LRSize \Gamma a \alpha$} \defas
  \forall \sigma : \Gamma' \leq \Gamma.\ \
  \RSize {\Gamma'} \alpha {a\sigma}
  % \exists a'.\
  % \RSize {\Gamma'} \alpha a' \mand
  % \Gamma' \der a\sigma = a' : \Size
\]

\fbox{$\LRT \Gamma {T'} {A'} \ell$} and \fbox{$\LR \Gamma t {T'} f {A'}$} defined by induction on ${A'} \in \SET[\ell]$.

@@

\nextcase $A' \evalsto N$ neutral.

  \noindent
  $\LRT{\Gamma}{T'}{A'}\ell$ iff
  $\forall \sigma : \Gamma' \leq \Gamma.\ \ \exists M.\ \ \RTy{\Gamma'}{A'}M \mand \Gamma' \der T'\sigma = M : \Set[\ell]$.

  \noindent
  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff
  $\forall \sigma : \Gamma' \leq \Gamma.\ \
   \exists m. \ \
   \R {\Gamma'} {\down {A'} f} m \mand
   \Gamma' \der t\sigma = m : T'\sigma$.

\nextcase $A' \evalsto \Nat \alpha$.

@@

  \noindent
  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff $\resurrect\Gamma \der T' = \Nat a$ for some $a$ and
  $\LRSize {\resurrect\Gamma} a \alpha$ and \\
  $\forall \sigma : \Gamma' \leq \Gamma.\ \
   \exists v. \ \
   \R {\Gamma'} {\down {A'} f} v \mand
   \Gamma' \der t\sigma = v : \Nat\,a\sigma$.

\nextcase $A' \evalsto \Set[\ell']$.

@@

\nextcase $A' \evalsto \piT A B$.

\noindent
$\LRT \Gamma {T'} {A'} \ell$ iff
\[
\begin{array}{l}
\Gamma \der T' = \piT U T : \Set[\ell] \mforsome U, T \mand \LRT \Gamma U A \ell \mand \\
\forall \sigma : \Gamma' \leq \Gamma.\ \
\LR {\Gamma'} u {U\sigma} a A \implies \LRT {\Gamma'} {T(\sigma,u)} {B[a]} \ell
. \\
\end{array}
\]
$\LR \Gamma t {T'} f {A'}$ iff
\[
\begin{array}{l}
\resurrect\Gamma \der T' = \piT U T : \Set[\ell] \mforsome U, T \mand \LRT {\resurrect\Gamma} U A \ell \mand \\
\forall \sigma : \Gamma' \leq \Gamma.\ \
\LR {\Gamma'} u {U\sigma} a A \implies \LR {\Gamma'} {t\sigma\,u} {T(\sigma,u)} {f\,a} {B[a]}
.
\end{array}
\]

\nextcase $A' \evalsto \piT \Size B$.

\noindent
$\LRT \Gamma {T'} {A'} \ell$ iff
\[
\begin{array}{l}
\Gamma \der T' = \piT \Size T : \Set[\ell] \mforsome T\\
\forall \sigma : \Gamma' \leq \Gamma.\ \
\LRSize {\Gamma'} a \alpha
\implies \LRT {\Gamma'} {T(\sigma,a)} {B[\alpha]} \ell
.
\end{array}
\]
$\LR \Gamma t {T'} f {A'}$ iff
\[
\begin{array}{l}
\resurrect\Gamma \der T' = \piT \Size T : \Set[\ell] \mforsome T\\
\forall \sigma : \Gamma' \leq \Gamma.\ \
\LRSize {\Gamma'} a \alpha
\implies \LR {\Gamma'} {t\sigma\,a} {T(\sigma,a)} {f\,\alpha} {B[\alpha]}
.
\end{array}
\]

\nextcase $A' \evalsto \forallT B$.

\noindent
$\LRT \Gamma {T'} {A'} \ell$ iff
\[
\begin{array}{l}
\Gamma \der T' = \forallT T : \Set[\ell] \mforsome T\\
\forall \sigma : \Gamma' \leq \Gamma,\
{\Gamma'} \der b : \Size,\
\beta \in \SIZE. \ \
\LRSize {\Gamma'} b \beta
\implies \LRT {\Gamma'} {T(\sigma,b)} {B[\beta]} \ell
.
\end{array}
\]
$\LR \Gamma t {T'} f {A'}$ iff
\[
\begin{array}{l}
\resurrect\Gamma \der T' = \forallT T : \Set[\ell] \mforsome T\\
\forall \sigma : \Gamma' \leq \Gamma,\
\Gamma' \der a : \Size,\
\resurrect{\Gamma'} \der b : \Size,\
\alpha,\beta \in \SIZE. \ \\ \qquad
\LRSize {\resurrect{\Gamma'}} b \beta
\implies \LR {\Gamma'} {t\sigma\,\ann a} {T(\sigma,b)} {f\,\ann\alpha} {B[\beta]}
.
\end{array}
\]

@@

\begin{theorem}[Into and out of the logical relation]
  \bla
  Let $A \in \Set[\ell]$.
  \begin{enumerate}

  \item $\LR \Gamma t T {\up A n} A$ if for all $\sigma : \Gamma' \leq \Gamma$ there is an $m$ such that
     $\RNe{\Gamma'}n m$ and $\Gamma' \der t\sigma = m : T$.

  \item If\/ $\LRT{\Gamma} T A \ell$ then for all $\sigma : \Gamma' \leq \Gamma$ there is some normal form $V$ such that
  $\Rty{\Gamma'} A V$ and $\Gamma' \der T\sigma = V : \Set[\ell]$.

  \item If\/ $\LR \Gamma t T f A$ then for all $\sigma : \Gamma' \leq \Gamma$ there is some normal form $v$ such that
  $\R{\Gamma'} {\down A f} v$ and $\Gamma' \der t\sigma = v : T\sigma$.

@@

Logical relation for substitutions.
\fbox{$\LRSub \Gamma \sigma \tau \Delta \eta \rho$}.
\begin{gather*}
  \ru{\der \Gamma
    }{\LRSub \Gamma {\sempty} {\sempty} {\cempty} {\sempty} {\sempty}}
\\[2ex]
  \ru{\LRSub \Gamma \sigma \tau \Delta \eta \rho \qquad
      \resurrect \Delta \der T \qquad
      \Gamma \der u = t : T\tau \qquad
      \LR \Gamma t {T \tau} f {T \rho} \qquad
      f = g \in T\rho
    }{\LRSub \Gamma {(\sigma,u)} {(\tau,t)} {\cext \Delta T} {(\eta,f)} {(\rho,q)}}
\\[2ex]
  \ru{\LRSub \Gamma \sigma \tau \Delta \eta \rho \qquad
      \Gamma \der a : \Size \qquad
      \LRSize \Gamma a \alpha
    }{\LRSub \Gamma {(\sigma,a)} {(\tau,a)} {\cext \Delta \Size} {(\eta,\alpha)} {(\rho,\alpha)}}
\\[2ex]
  \ru{\LRSub \Gamma \sigma \tau \Delta \eta \rho \qquad
      \Gamma \der a, b : \Size \qquad
      \alpha,\beta \in \SIZE \qquad
      \LRSize \Gamma b \beta
    }{\LRSub \Gamma {(\sigma,a)} {(\tau,b)} {\erext \Delta \Size} {(\eta,\alpha)} {(\rho,\beta)}}
\end{gather*}

>>>>

\fbox{$\Gamma \der a : \Size$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der a : \Size$).
\[
  \ru{\der \Gamma}{\Gamma \der \infty : \Size}
\qquad
  \rux{\der \Gamma}{\Gamma \der o : \Size}{o \in \NN}
\qquad
  \ru{\der \Gamma \qquad \Gamma(i) = \erann[\noterased]\Size
    }{\Gamma \der \ind i + o : \Size
    }
\]

@@

      % \Gamma \der a : \Size \qquad
      \resurrect \Gamma \der a, b : \Size

@@

  \ru{\resurrect\Gamma \der a, b : \Size \qquad

@@

      % \Gamma \der a : \Size \qquad

@@

    }{\Gamma \der (\zero{a})\,\case[\ell] T {t_z} {t_s} = t_z : T\,\zero{b}}

@@

  \ru{\resurrect\Gamma \der a : \Size \qquad
      \Gamma \der t : \Nat b \qquad

@@

@@

    }{\Gamma \der (\suc{a} t)\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,(\suc{b} t)}

@@

      \resurrect\Gamma \der a : \Size \qquad

@@

  \ru{\Gamma \der t : \forallT T \qquad
      \cext {\resurrect\Gamma} \Size \der a : \Size
      %\erext \Gamma \Size \der a : \Size

@@

      % \Gamma \der a, a' : \Size \qquad
      \resurrect\Gamma \der a,a',b : \Size

@@

  \ru{%\Gamma \der a, a' : \Size \qquad
      \resurrect\Gamma \der a, a', b : \Size

@@

  \ru{\resurrect \Gamma \der a, a' : \Size \qquad

@@

Substitution typing. % and equality.
\fbox{$\Gamma \der \tau : \Delta$}
(implies $\der \Gamma$ and $\der \Delta$
 %% and $\Gamma \der \tau : \resurrect \Delta$  %NO
 [and $\resurrect\Gamma \der \tau : \Delta$]
 and $\resurrect\Gamma \der \tau : \resurrect \Delta$).

\begin{gather*}
  \ru{\der \Gamma \qquad
    }{\Gamma \der \sempty : \cempty}
\qquad
  \ru{\Gamma \der \tau : \Delta \qquad
      \resurrect \Delta \der T \qquad
      \Gamma \der t : T \tau
    }{\Gamma \der \sext \tau t : \cext \Delta T}
\qquad
  \ru{\Gamma \der \tau : \Delta \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \sext \tau a : \cext \Delta \Size}
\qquad
  \ru{\Gamma \der \tau : \Delta \qquad
      \resurrect\Gamma \der a : \Size
    }{\Gamma \der \sext \tau a : \erext \Delta \Size}
\end{gather*}

Liftings: $\fbox{$\xi : \Gamma \leq \Delta$}
  \defas \Gamma \der \xi : \Delta \mand \xi = \slift^k_m \mforsome k,m$
%  \defas \xi = \slift^k_m \mwith \Gamma \der \xi : \Delta$
(implies $\der \Gamma$ and $\der \Delta$).
% (implies $\der \Gamma$ and $\der \Delta$ and $\Gamma \der \xi : \Delta$).
% \begin{gather*}
%   \ru{\der \Gamma
%     }{\sempty : \Gamma \leq \cempty}
% \qquad
%   \ru{\xi : \Gamma \leq \Delta \qquad
%       \resurrect\Gamma \der T
%     }{\xi \slift : \cext \Gamma T \leq \Delta}
% \qquad
%   \ru{\xi : \Gamma \leq \Delta \qquad
%       \resurrect \Delta \der U \qquad
%       \resurrect\Gamma \der T \leq U \xi
%     }{(\xi \slift, \ind 0) : \cext \Gamma T \leq \cext \Delta U}
% \end{gather*}

@@

\begin{gather*}
  \ru{\der \Gamma \qquad
    }{\sueq \Gamma \sempty \sempty \sempty \cempty}
\qquad
  \ru{\sueq \Gamma \sigma {\sigma'} \tau \Delta \qquad
      \resurrect \Delta \der T \qquad
      \Gamma \der u = u' = t : T \tau
    }{\sueq \Gamma {\sext \sigma u } {\sext {\sigma'} {u'}} {\sext \tau t} {\cext \Delta T}}
\\[2ex]
  \ru{\sueq \Gamma \sigma {\sigma'} \tau \Delta \qquad
      \Gamma \der a = a' = b : T \tau
    }{\sueq \Gamma {\sext \sigma a } {\sext {\sigma'} {a'}} {\sext \tau b} {\cext \Delta \Size}}
\qquad
  \ru{\sueq \Gamma \sigma {\sigma'} \tau \Delta \qquad
      \resurrect\Gamma \der a, a', b : T \tau
    }{\sueq \Gamma {\sext \sigma a } {\sext {\sigma'} {a'}} {\sext \tau b} {\erext \Delta \Size}}
\end{gather*}

@@

\begin{lemma}[Resurrection]
\label{lem:res} \bla
\begin{enumerate}
\item $\der \Gamma$ iff $\der \resurrect\Gamma$;  then $\resurrect \Gamma \der \sid : \Gamma$.
  % Consequently, if\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
  % Further, if $\Gamma \der J : \resurrect\Delta$ then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
\item If\/ $\Gamma \der J : \resurrect \Delta$ % and $\der \Delta$
  then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der \sigma : \Delta$ % then $\Gamma \der \sigma : \resurrect\Delta$.
  then $\resurrect\Gamma \der \sigma : \resurrect\Delta$.
\end{enumerate}
\end{lemma}

@@

\begin{lemma}[Substitution equality]
\label{lem:subeq}\bla
\begin{enumerate}
\item Conversion: If\/ $\sueq \Gamma \sigma {\sigma'} {\tau_1} \Delta$ and $\sueq {\resurrect\Gamma} {\tau_1} {\tau_2} {\tau} {\resurrect\Delta}$ then $\sueq \Gamma \sigma {\sigma'} {\tau_2} \Delta$.
\item Reflexivity: If\/ $\Gamma \der \sigma : \Delta$ then $\sueq \Gamma \sigma \sigma \sigma \Delta$.
\item Symmetry:  If\/ $\sueq \Gamma \sigma {\sigma'} {\tau} \Delta$ then $\sueq \Gamma {\sigma'} {\sigma} {\tau} \Delta$.
\item Transitivity:  If\/ $\sueq \Gamma {\sigma_1} {\sigma_2} {\tau} \Delta$ and $\sueq \Gamma {\sigma_2} {\sigma_3} {\tau} \Delta$ then $\sueq \Gamma {\sigma_1} {\sigma_3} {\tau} \Delta$.
\item Functionality: Let $\sueq \Gamma \sigma {\sigma'} \tau \Delta$.
  \begin{enumerate}
  \item
  If $\Delta \der t : T$ then $\Gamma \der t\sigma = t\sigma' : T\tau$.
  \item
  If $\Delta \der t = t' : T$ then $\Gamma \der t\sigma = t'\sigma' : T\tau$.
  \item
  Corollary: If $\Delta \der T \leq T'$ then $\Gamma \der T\sigma \leq T\sigma'$.
  \end{enumerate}
\end{enumerate}
\end{lemma}
\begin{corollary}[Partial resurrection for substitution equality]
  If\/ $\sueq \Gamma \sigma {\sigma'} {\tau} \Delta$ then
  $\sueq {\resurrect \Gamma} \tau \tau \tau {\resurrect \Delta}$.
\end{corollary}

@@

\begin{theorem}[Reflection and reification]
\label{thm:rere}
Let $T \in \SET[\ell]$ and $T \shape S_1$ and $T \shape S_2$.
\begin{enumerate}
\item If $n_1 = n_2 \in \NE$ then $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$.
\item If $t_1 = t_2 \in T$ then $\down{S_1}{t_1} = \down{S_2}{t_2} \in \NF$.
\end{enumerate}
\end{theorem}
\begin{proof}
By induction on $T \in \SET[\ell]$ and cases on $T \shape S_1$ and $T \shape S_2$.
\begin{caselist}

@@

\begin{corollary}
  \label{cor:rere}
  Let\/ $T \in \SET[\ell]$.
  \begin{enumerate}
  \item If $n = n' \in \NE$ then $\up T n = \up T n' \in T$.
  \item If $t = t' \in T$ then $\down T t = \down T {t'} \in \NF$.
  \end{enumerate}
\end{corollary}

@@

\begin{theorem}[Fundamental theorem]
  \label{thm:fund}
  \bla
  \begin{enumerate}
  \item
  If $\der \Gamma$ then $\models \Gamma$.
  \item
  If $\Gamma \der J$ then % $\models \Gamma$ and
  $\Gamma \models J$
  \end{enumerate}
\end{theorem}
\begin{proof}
Simultaneously, by induction on the derivation.
% The soundness of subtyping and the conversion rule is due to subsumption (lemmata \ref{lem:subsump} and \ref{lem:usubsump}).

@@

We write $\var\Gamma$ for $\var{|\Gamma|}$.
We define the identity environment $\rho_\Gamma$ by induction on $\Gamma$.
We set $\rho_{\cempty} = \sempty$ and
\[
\begin{array}{lll}
  \rho_{\cext \Gamma T} & = & (\rho_\Gamma, \up{T\rho_\Gamma} \var\Gamma) \\
  \rho_{\eext \Gamma \Size} & = & (\rho_\Gamma, \var\Gamma) \\
\end{array}
\]
\begin{lemma}[Identity environment]
  If $\der \Gamma$ then
  $\rho_\Gamma \in \Gamma$.
  % $\rho_\Gamma = \rho_\Gamma \eeq \rho_\Gamma \in \Gamma$
\end{lemma}
\begin{proof}
By induction on $\der \Gamma$.
\begin{caselist}

\nextcase
\[
  \ru{\der \Gamma \qquad \resurrect\Gamma \der T
    }{\der \cext\Gamma T}
\]
By induction hypothesis $\rho_\Gamma \in \Gamma$.
By resurrection (Lemma~\ref{lem:resenv}) $\rho_\Gamma \in \resurrect\Gamma$.
By the fundamental theorem (Thm.~\ref{thm:fund}) we have $A := T\rho_\Gamma \in \SET[\ell]$ for some $\ell$.
By reflection (Cor.~\ref{cor:rere}) it follows that $\up A \var\Gamma \in A$, thus
$(\rho_\Gamma, \up A \var\Gamma) \in \cext \Gamma T$.

\nextcase
\[
  \ru{\der \Gamma
    }{\der \eext\Gamma \Size}
\]
By induction hypothesis $\rho_\Gamma \in \Gamma$.  Since $\var\Gamma \in \SIZE$ we conclude
$(\rho_\Gamma, \var\Gamma) \in \eext\Gamma \Size)$.
\end{caselist}
\end{proof}

We can now define the normalization relation
$\nbe \Gamma T t \evalsto v \iff \R {|\Gamma|} {\down{T \rho_\Gamma} (t \rho_\Gamma)} v$.

\begin{theorem}[Completeness of NbE]
  If\/ $\Gamma \der t = t' : T$ then there are normal forms $v \approx v'$ such that
  $\nbe \Gamma T t \evalsto v$ and $\nbe \Gamma T {t'} \evalsto v'$.
\end{theorem}
\begin{proof}
  By the fundamental theorem, $T \rho_\Gamma \in \SET[\ell]$ for some $\ell$ and
  $t \rhog = t' \rhog \in T\rhog$.  By reification (Cor.~\ref{cor:rere}) we have
  $\down{T\rhog} (t\rhog) = \down{T\rhog}(t'\rhog) \in \NF$ which implies the theorem
  by read back with $k = |\Gamma|$.
\end{proof}

@@

\[
  \fbox{$\LRSize \Gamma a \alpha$} \defas
  \forall \xi : \Gamma' \leq \Gamma.\ \
  \RSize {\Gamma'} \alpha {a\xi}
  % \exists a'.\
  % \RSize {\Gamma'} \alpha a' \mand
  % \Gamma' \der a\xi = a' : \Size
\]

We simultaneously define
\fbox{$\LRT \Gamma {T'} {A'} \ell$} for $\Gamma \der T' : \Set[\ell]$
and \fbox{$\LR \Gamma t {T'} f {A'}$} for $\Gamma \der t : T'$ and $f \in A'$
defined by induction on ${A'} \in \SET[\ell]$.

@@

\vspace{2ex}
\nextcase $A' \evalsto N$ neutral.

  \noindent
  $\LRT{\Gamma}{T'}{A'}\ell$ iff
  $\forall \xi : \Gamma' \leq \Gamma.\ \ \exists M.\ \ \RTy{\Gamma'}{A'}M \mand \Gamma' \der T'\xi = M : \Set[\ell]$.

  \noindent
  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff
  $\forall \xi : \Gamma' \leq \Gamma.\ \
   \exists m. \ \
   \R {\Gamma'} {\down {A'} f} m \mand
   \Gamma' \der t\xi = m : T'\xi$.

\vspace{2ex}
\nextcase $A' \evalsto \Nat \alpha$.

@@

  \noindent
  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff $\resurrect\Gamma \der T' = \Nat a$ for some $a$ and
  $\LRSize {\resurrect\Gamma} a \alpha$ and \\
  $\forall \xi : \Gamma' \leq \Gamma.\ \
   \exists v. \ \
   \R {\Gamma'} {\down {A'} f} v \mand
   \Gamma' \der t\xi = v : \Nat\,a\xi$.

\vspace{2ex}
\nextcase $A' \evalsto \Set[\ell']$.

@@

\vspace{2ex}
\nextcase $A' \evalsto \piT A B$.

\noindent
$\LRT \Gamma {T'} {A'} \ell$ iff
\[
\begin{array}{l}
\Gamma \der T' = \piT U T : \Set[\ell] \mforsome U, T \mand \LRT \Gamma U A \ell \mand \\
\forall \xi : \Gamma' \leq \Gamma.\ \
\LR {\Gamma'} u {U\xi} a A \implies \LRT {\Gamma'} {T(\xi,u)} {B[a]} \ell
. \\
\end{array}
\]
$\LR \Gamma t {T'} f {A'}$ iff
\[
\begin{array}{l}
\resurrect\Gamma \der T' = \piT U T : \Set[\ell] \mforsome U, T \mand \LRT {\resurrect\Gamma} U A \ell \mand \\
\forall \xi : \Gamma' \leq \Gamma.\ \
\LR {\Gamma'} u {U\xi} a A \implies \LR {\Gamma'} {t\xi\,u} {T(\xi,u)} {f\,a} {B[a]}
.
\end{array}
\]

\vspace{2ex}
\nextcase $A' \evalsto \piT \Size B$.

\noindent
$\LRT \Gamma {T'} {A'} \ell$ iff
\[
\begin{array}{l}
\Gamma \der T' = \piT \Size T : \Set[\ell] \mforsome T\\
\forall \xi : \Gamma' \leq \Gamma.\ \
\LRSize {\Gamma'} a \alpha
\implies \LRT {\Gamma'} {T(\xi,a)} {B[\alpha]} \ell
.
\end{array}
\]
$\LR \Gamma t {T'} f {A'}$ iff
\[
\begin{array}{l}
\resurrect\Gamma \der T' = \piT \Size T : \Set[\ell] \mforsome T\\
\forall \xi : \Gamma' \leq \Gamma.\ \
\LRSize {\Gamma'} a \alpha
\implies \LR {\Gamma'} {t\xi\,a} {T(\xi,a)} {f\,\alpha} {B[\alpha]}
.
\end{array}
\]

\vspace{2ex}
\nextcase $A' \evalsto \forallT B$.

\noindent
$\LRT \Gamma {T'} {A'} \ell$ iff
\[
\begin{array}{l}
\Gamma \der T' = \forallT T : \Set[\ell] \mforsome T\\
\forall \xi : \Gamma' \leq \Gamma,\
{\Gamma'} \der b : \Size,\
\beta \in \SIZE. \ \
\LRSize {\Gamma'} b \beta
\implies \LRT {\Gamma'} {T(\xi,b)} {B[\beta]} \ell
.
\end{array}
\]
$\LR \Gamma t {T'} f {A'}$ iff
\[
\begin{array}{l}
\resurrect\Gamma \der T' = \forallT T : \Set[\ell] \mforsome T\\
\forall \xi : \Gamma' \leq \Gamma,\
%\Gamma' \der a : \Size,\
\resurrect{\Gamma'} \der a,b : \Size,\
\alpha,\beta \in \SIZE. \ \\ \qquad
\LRSize {\resurrect{\Gamma'}} b \beta
\implies \LR {\Gamma'} {t\xi\,\ann a} {T(\xi,b)} {f\,\ann\alpha} {B[\beta]}
.
\end{array}
\]

@@

\begin{lemma}[Weakening]
  Let $\xi : \Gamma' \leq \Gamma$.
  \begin{enumerate}
  \item If\/ $\LRSize \Gamma a \alpha$ then $\LRSize{\Gamma'}{a\xi}\alpha$.
  \item If\/ $\LRT {\Gamma} T A \ell$ then $\LRT{\Gamma'}{T\xi}A \ell$.
  \item If\/ $\LR \Gamma t T f A$ then $\LR{\Gamma'}{t\xi}{T\xi} f A$.
  \end{enumerate}
\end{lemma}

\begin{theorem}[Into and out of the logical relation]
  \label{thm:into}
  \bla
  Let $A \in \SET[\ell]$.
  \begin{enumerate}

  \item \label{it:into}
     $\LR \Gamma t T {\up A n} A$ if for all $\xi : \Gamma' \leq \Gamma$ there is an $m$ such that
     $\RNe{\Gamma'}n m$ and\/ $\Gamma' \der t\xi = m : T$.

  \item \label{it:outty}
     If\/ $\LRT{\Gamma} T A \ell$ then for all $\xi : \Gamma' \leq \Gamma$
     there is some normal form $V$ such that
     $\Rty{\Gamma'} A V$ and $\Gamma' \der T\xi = V : \Set[\ell]$.

  \item \label{it:out}
    If\/ $\LR \Gamma t T f A$ then for all $\xi : \Gamma' \leq \Gamma$
    there is some normal form $v$ such that
    $\R{\Gamma'} {\down A f} v$ and $\Gamma' \der t\xi = v : T\xi$.

@@

\begin{proof}
We prove the propositions for $A'$ and $T'$ (instead of $A$ and $T$)
simultaneously by induction on $A' \in \SET[\ell]$.

\nextcase $A' \evalsto \piT A B$ and $\resurrect\Gamma \der T' = \piT U T$.
\begin{enumerate}
\item To demonstrate $\LR \Gamma t {T'} {\up {A'} n} {A'}$ we assume
$\xi : \Gamma' \leq \Gamma$ and $\LR{\Gamma'}{u}{U\xi} a A$ and show
$\LR{\Gamma'} {(t\xi\,u)} {T(\xi,u)} {\up{B[a]}(n\,\down A a)} {B[a]}$.
This follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$ if
we can prove that for all $\xi' : \Gamma'' \leq \Gamma'$ there is a neutral normal form $m$
such that $\RNe{\Gamma''} {(n\,\down A a)} m$ and $\Gamma'' \der (t\xi\,u)\xi' = m : T(\xi,u)\xi'$.
Note that the last goal can be rewritten to $\Gamma'' \der t\xi\xi'\,(u\xi') = m : T(\xi\xi',u\xi')$.
We know that $\RNe{\Gamma''} n {m_0}$ with $\Gamma'' \der t\xi\xi' = m_0 : T'\xi\xi'$.
Thus, it is sufficient to show that there is some normal form $v$ such that
$\R{\Gamma''} {\down A a} v$ and $\Gamma'' \der u\xi' = v : U\xi\xi'$.
However, this follows by induction hypothesis (\ref{it:out}) on
assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.
\end{enumerate}
\end{caselist}
\end{proof}
